Altering Backward Pass Gradients to improve Convergence

### 2022-11-14 0100.tex
Added PAKDD version.

\documentclass[times,sort&compress]{elsarticle}
\usepackage{ycviu}
\usepackage{framed}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{amsthm}
\usepackage{pifont}
\usepackage{graphicx}

% \usepackage{xcolor} \usepackage{pagecolor} \pagecolor{black} \color{white}

\graphicspath{{images/}}

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\NewDocumentCommand{\codeword}{v}{\texttt{\textcolor{orange}{#1}}}
\newcommand{\cmark}{\ding{51}} \newcommand{\xmark}{\ding{55}}

\definecolor{newcolor}{rgb}{.8,.349,.1}
\definecolor{codegreen}{rgb}{0.7,0.7,0.7}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{1,1,1}
\lstdefinestyle{mystyle}{ backgroundcolor=\color{backcolour},   
commentstyle=\color{codegreen}, keywordstyle=\color{magenta},
numberstyle=\tiny\color{codegray}, stringstyle=\color{codepurple},
basicstyle=\ttfamily\footnotesize, breakatwhitespace=false,         
breaklines=true,                 
captionpos=b,                    
keepspaces=true,                 
numbersep=5pt,                  
showspaces=false,                
showstringspaces=false, showtabs=false,                  
tabsize=2 }
\lstset{style=mystyle}

\journal{Computer Vision and Image Understanding}

\begin{document}

\thispagestyle{empty}

\begin{table*}[!th]

\begin{minipage}{.9\textwidth}
\baselineskip12pt
\ifpreprint
  \vspace*{1pc}
\else
  \vspace*{-6pc}
\fi

\noindent {\LARGE\itshape Computer Vision and Image Understanding}
\vskip6pt

\noindent {\Large\bfseries Authorship Confirmation}

\vskip1pc

{\bf Please save a copy of this file, complete and upload as the ``Confirmation of
Authorship'' file.}

\vskip1pc

As corresponding author I, \underline{\hphantom{\hspace*{7cm}}}, hereby confirm on
behalf of all authors that:

\vskip1pc

\begin{enumerate}
\itemsep=3pt
\item This manuscript, or a large part of it, \underline {has not been published,  was
not, and is not being submitted to} any other journal.

\item If \underline {presented} at or \underline {submitted} to or \underline {published
}at a conference(s), the conference(s) is (are) identified and  substantial \underline
{justification for re-publication} is presented  below. A \underline {copy of conference
paper(s) }is(are) uploaded with the  manuscript.

\item If the manuscript appears as a preprint anywhere on the web, e.g. arXiv,  etc., it
is identified below. The \underline {preprint should include a  statement that the paper
is under consideration at Computer Vision and Image Understanding}.

\item All text and graphics, except for those marked with sources, are \underline
{original works} of the authors, and all necessary permissions for  publication were
secured prior to submission of the manuscript.

\item All authors each made a significant contribution to the research reported  and
have \underline {read} and \underline {approved} the submitted  manuscript. 
\end{enumerate}

Signature\underline{\hphantom{\hspace*{7cm}}} Date\underline{\hphantom{\hspace*{4cm}}}
\vskip1pc

\rule{\textwidth}{2pt}
\vskip1pc

{\bf List any pre-prints:} \vskip5pc

\rule{\textwidth}{2pt}
\vskip1pc

{\bf Relevant Conference publication(s) (submitted, accepted, or published):} \vskip5pc

{\bf Justification for re-publication:}

\end{minipage}
\end{table*}

\clearpage
\thispagestyle{empty}
\ifpreprint
  \vspace*{-1pc}
\fi

\begin{table*}[!t]
\ifpreprint\else\vspace*{-15pc}\fi

\section*{Research Highlights (Required)}

To create your highlights, please type the highlights against each \verb+\item+ command.

\vskip1pc

\fboxsep=6pt \fbox{
\begin{minipage}{.95\textwidth}
It should be short collection of bullet points that convey the core findings of the
article. It should  include 3 to 5 bullet points (maximum 85 characters, including
spaces, per bullet point.)  
\vskip1pc
\begin{itemize}

\item Altering gradients in the backward pass can significantly alter a CNN's
performance.

\item PowerGrad Transform is proposed that enhances the backward pass gradients.

\item Theoretical analysis of the impact of PowerGrad transformation is conducted.

\item Efficacy of PowerGrad Transform is demonstrated through experiments.

\end{itemize}
\vskip1pc
\end{minipage}
}

\end{table*}

\ifpreprint
  \setcounter{page}{1}
\else
  \setcounter{page}{1}
\fi

\begin{frontmatter}

\title{Altering Backward Pass Gradients to improve Convergence}

\author[1]{Bishshoy Das\corref{cor1}}
\ead{bishshoy.das@ee.iitd.ac.in}

\author[1]{Milton Mondal\corref{cor2}}
\ead{milton.mondal@ee.iitd.ac.in}

\author[1]{\\ Brejesh Lall}
\ead{brejesh@ee.iitd.ac.in}

\author[1]{Shiv Dutt Joshi}
\ead{sdjoshi@ee.iitd.ac.in}

\author[1]{Sumantra Dutta Roy}
\ead{sumantra@ee.iitd.ac.in}

\cortext[cor1]{Corresponding author}
\cortext[cor2]{Equal contribution}

\address[1]{Electrical Engineering Department, Indian Institute of Technology Delhi, Hauz Khas, New Delhi - 110016, India}

\received{1 May 2013}
\finalform{10 May 2013}
\accepted{13 May 2013}
\availableonline{15 May 2013}
\communicated{S. Sarkar}

\begin{abstract}
In standard neural network training, the gradients in the backward pass are determined
by the forward pass. As a result, the two stages are coupled. This is how most neural
networks are trained currently. However, gradient modification in the backward pass has
seldom been studied in the literature. In this paper we explore decoupled training,
where we alter the gradients in the backward pass. We propose a simple yet powerful
method called PowerGrad Transform, that alters the gradients before the weight update in
the backward pass and significantly enhances the predictive performance of the neural
network. PowerGrad Transform trains the network to arrive at a better optima at
convergence. It is computationally extremely efficient, virtually adding no additional
cost to either memory or compute, but results in improved final accuracies on both the
training and test sets. PowerGrad Transform is easy to integrate into existing training
routines, requiring just a few lines of code. PowerGrad Transform accelerates training
and makes it possible for the network to better fit the training data. With decoupled
training, PowerGrad Transform improves baseline accuracies for ResNet-50 by 0.73\%, for
SE-ResNet-50 by 0.66\% and by more than 1.0\% for the non-normalized ResNet-18 network
on the ImageNet classification task.
\end{abstract}

\begin{keyword}
\KWD backpropagation \sep gradients \sep neural networks \sep softmax \sep clipping
\end{keyword}

\end{frontmatter}

@@

\begin{equation}
\hspace{4cm}
W_i^{t+1} = W_i^t - \lambda\ \nabla_{W_i}(L)
\hspace{2cm}
i=D,D-1,\ldots,1
\end{equation}

@@

\begin{equation}
\hspace{4cm}
W_i^{t+1} = W_i^t - \lambda\ f(\nabla_{W_i}(L))
\hspace{2cm}
i=D,D-1,\ldots,1
\end{equation}

@@

\begin{align}
\hspace{4cm}
W_D^{t+1} &= W_D^t - \lambda\ f(\nabla_{W_D}(L)) & \\
W_i^{t+1} &= W_i^t - \lambda\ \nabla_{W_i}(L)^{*} &\hspace{-3cm} i=D-1,\ldots,1
\end{align}

@@

Gradient Clipping (GC): Gradient Clipping \cite{pascanu2013difficulty}, often used in
natural language processing methods \cite{merity2017regularizing}, is a technique that
involves changing or clipping the gradients with respect to a predefined threshold value
during backward propagation through the network and updating the weights using the
clipped gradients \cite{zhang2019gradient, smith2020generalization}. By rescaling the
gradients, the weight updates are likewise rescaled, significantly reducing the risk of
an overflow or underflow \cite{pascanu2012understanding}. GC can be used for training
networks without batch-normalization.

@@

Knowledge Distillation: The formulation of PowerGrad Transform is akin to temperature
based logit suppression of \cite{hinton2015distilling}. Knowledge distillation
\cite{DBLP:conf/nips/BaC14} is a process in which two networks are trained. First a
teacher network is trained on a given dataset and then the soft-labels (the predicted
probabilities) of the teacher network are used to train another network, called the
student network. In similar style of label smoothing, knowledge distillation aims to
generate smooth gradient behaviour by forcing the network not to overpredict on any
single image. As in case of label smoothing, the student network's weights are
automatically penalized if the network assigns a probability value higher than the soft
labels generated by the teacher network. Variants of knowledge distillation include
self-distillation \cite{zhang2019your, wang2021memory}, identical student network
distillation \cite{furlanello2018born}, channel distillation \cite{ge2019distilling},
regularizing wrong predictions \cite{yun2020regularizing} and representation or
embedding based knowledge distillation \cite{aguilar2020knowledge, yao2018deep,
passalis2018unsupervised}. Distillation is applied to various tasks such as model
compression \cite{wang2020real}, face recognition \cite{ge2018low}, enhancing ensemble
performance \cite{zhang2018deep}, interpreting neural networks \cite{liu2018improving}.
Efficacy of distillation is extensivly studied in \cite{cho2019efficacy,
yuan2020revisiting}. Applications of logit suppression include: deep metric learning
\cite{zhai2018classification}, non-parametric neighborhood component analysis
\cite{wu2018improving}, sequence-to-sequence modelling \cite{chorowski2016towards},
reinforcement learning \cite{he2018determining} and face recognition
\cite{liu2017sphereface, wang2018cosface}.

@@

\begin{proof}[\unskip\nopunct] \begin{alignat}{2} &\ \ \frac{p_i^{\alpha }}{\sum
_{j=1}^C p_j^{\alpha}} < p_i \\ \Rightarrow &\ p_i^{\alpha -1}<\sum _{j=1}^C p_j^{\alpha
} \\ \Rightarrow &\ p_i > \Big(\sum _{j=1}^C p_j^{\alpha}\Big)^{\frac{1}{\alpha-1}}
\end{alignat} \end{proof}

@@

\begin{proof}[\unskip\nopunct] \begin{equation} \frac{\partial}{\partial \alpha
}\left(\sum _{j=1}^c p_j^{\alpha}\right){}^{\frac{1}{\alpha -1}}\ \ \geq \ \ 0
\end{equation} \end{proof}

@@

\begin{table}[!t]
\centering
\caption{ Results and comparison table for CNN networks trained on ImageNet-1K (sorted
in descending order w.r.t. the last column denoting test accuracy improvements over
baseline). Best training and test accuracies are highlighted in red and blue
respectively. Accuracy differences are highlighted in yellow. In all experiments, we
observe that PGT positively improves both training and test accuracies. We also observe
that deeper networks such as SE-ResNet-50 and ResNet-50 accrue higher gains over the
baselines as compared to shallow networks like ResNet-18. }
\label{tab:imagenet_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{Model}} & \multirow{2}{*}{\textbf{Scheduler}} &
\multirow{2}{*}{\textbf{Method}} & \textbf{PowerGrad} & \textbf{\hspace{-0.65cm} Train}
& \textbf{Train} & \textbf{\hspace{-0.65cm} Test} & \textbf{Test} \\
& & & \textbf{Transform ($\alpha$)} & \textbf{Accuracy (\%)} & \textbf{Diff (\%)} &
\textbf{Accuracy (\%)} & \textbf{Diff (\%)} \\
\midrule
\multirow{2}{*}{SE-ResNet-50} & \multirow{2}{*}{Cosine} & Baseline & - & 81.5 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.97}}} & 77.218 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.734}}} \\
& & PGT & 0.3 & \textcolor{red}{\textbf{82.47}} & & \textcolor{blue}{\textbf{77.952}} &
\\
\midrule
\multirow{2}{*}{ResNet-50} & \multirow{2}{*}{Cosine} & Baseline & - & 79.18 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.5}}} & 76.56 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.656}}} \\
& & PGT & 0.05 & \textcolor{red}{\textbf{79.68}} & & \textcolor{blue}{\textbf{77.216}} &
\\
\midrule
\multirow{2}{*}{ResNet-50} & \multirow{2}{*}{Step} & Baseline & - & 78.99 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.57}}} & 75.97 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.524}}} \\
& & PGT & 0.05 & \textcolor{red}{\textbf{79.56}} & & \textcolor{blue}{\textbf{76.494}} &
\\
\midrule
\multirow{2}{*}{ResNet-101} & \multirow{2}{*}{Cosine} & Baseline & - & 82.29 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.81}}} & 77.896 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.362}}} \\
& & PGT & 0.3 & \textcolor{red}{\textbf{83.1}} & & \textcolor{blue}{\textbf{78.258}} &
\\
\midrule
\multirow{2}{*}{SE-ResNet-18} & \multirow{2}{*}{Cosine} & Baseline & - & 71.42 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.18}}} & 71.09 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.346}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{71.6}} & & \textcolor{blue}{\textbf{71.436}} &
\\
\midrule
\multirow{2}{*}{ResNet-18} & \multirow{2}{*}{Step} & Baseline & - & 69.95 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.35}}} & 69.704 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.14}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{70.3}} & & \textcolor{blue}{\textbf{69.844}} &
\\
\midrule
\multirow{2}{*}{ResNet-18} & \multirow{2}{*}{Cosine} & Baseline & - & 70.38 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.15}}} & 70.208 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.09}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{70.53}} & & \textcolor{blue}{\textbf{70.298}} &
\\
\end{tabular}
\end{table}

@@

In our experiments with Squeeze-and-Excitation variant of ResNet-50 i.e.
SE-ResNet-50\cite{hu2018squeeze}, we observe significant improvements; a $0.97\%$ boost
in training accuracy and a $0.734\%$ increase in test accuracy. In our experiments with
ResNet-50, we find an increase of $0.5\%$ and $0.656\%$ performance enhancement over the
cosine scheduler baseline for training and testing respectively. The corresponding
improvement over the step scheduler baseline for ResNet-50 is $0.57\%$ (training) and
$0.524\%$ (testing). ResNet-101 sees a higher improvement in training fit, $0.81\%$ to
be exact, while the improvement over the test set is $0.362\%$. Smaller networks such as
SE-ResNet-18 and ResNet-18 sees accuracy boosts which are smaller but nevertheless
positive. Therefore we conclude that with PGT, the consistent improvements in training
accuracies across all cases is because the networks train better and arrive at better
optimas during convergence. Per epoch training and test accuracy plots of ResNet-18 and
ResNet-50 (both with and without PGT) are shown in Fig. \ref{fig:metrics}(a, b). For
better visualization of the top end of the plots, we scale both axes in log scale.
Because both training and test accuracies improve, it indicates that PowerGrad Transform
allows the network to learn better representations\footnote{Reproducible code, training
recipes for the experiments, pretrained checkpoints and training logs are provided at:
\\
\url{ https://github.com/bishshoy/power-grad-transform }. Code is adapted from
\cite{rw2019timm}.}.

@@

\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{baseline-w-layer-4-2}
\caption{Norm of filters of convolutional layer 11 (without PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{baseline-w-layer-7-2}
\caption{Norm of filters of the final convolutional layer (without PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{baseline-f-layer-22-1}
\caption{Norm of final layer's output features (without PGT)}
\end{subfigure}

\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{pgt-w-layer-4-2}
\caption{Norm of filters of convolutional layer 11 (with PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{pgt-w-layer-7-2}
\caption{Norm of filters of the final convolutional layer (with PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{pgt-f-layer-22-1}
\caption{Norm of final layer's output features (with PGT)}
\end{subfigure}
\captionsetup{font=normalsize}
\caption{ Experiment demonstrating the Zeroing Out phenomena, as observed in different
filters and feature vectors. PGT greatly reduces the zeroing out of filters and promote
better learning. }
\label{fig:norm_plots}
\end{figure}

@@

\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{baseline-w-layer-5-3}
\caption{Norm of filters of layer 15 (without AGC/PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{pgt-w-layer-5-3}
\caption{Norm of filters of layer 15 (with PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{agc_pgt-w-layer-5-3}
\caption{Norm of filters of layer 15 (with AGC+PGT)}
\end{subfigure}

\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{baseline-w-layer-1-2}
\caption{Norm of filters of layer 2 (without AGC/PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{pgt-w-layer-1-2}
\caption{Norm of filters of layer 2 (with PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.33\textwidth}
\includegraphics[width=\textwidth]{agc_pgt-w-layer-1-2}
\caption{Norm of filters of layer 2 (with AGC+PGT)}
\end{subfigure}
\captionsetup{font=normalsize}
\caption{ Experiment demonstrating the efficacy of PGT and AGC+PGT, as observed in
different layers. }
\label{fig:agc_pgt}
\end{figure}

@@

\bibliographystyle{model2-names}
\bibliography{references}

>>>>

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{framed}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
% \usepackage{amsthm}
\usepackage{pifont}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\graphicspath{{images/}}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\NewDocumentCommand{\codeword}{v}{\texttt{\textcolor{orange}{#1}}}
\newcommand{\cmark}{\ding{51}} \newcommand{\xmark}{\ding{55}}
\definecolor{newcolor}{rgb}{.8,.349,.1}
\definecolor{codegreen}{rgb}{0.7,0.7,0.7}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{1,1,1}
\lstdefinestyle{mystyle}{ backgroundcolor=\color{backcolour},   
commentstyle=\color{codegreen}, keywordstyle=\color{magenta},
numberstyle=\tiny\color{codegray}, stringstyle=\color{codepurple},
basicstyle=\ttfamily\footnotesize, breakatwhitespace=false,         
breaklines=true,                 
captionpos=b,                    
keepspaces=true,                 
numbersep=5pt,                  
showspaces=false,                
showstringspaces=false, showtabs=false,                  
tabsize=2 }
\lstset{style=mystyle}

\begin{document}
\title{Altering Backward Pass Gradients to improve Convergence}
\author{Bishshoy Das\inst{1}\orcidID{0000-0002-6525-4015}\and Milton Mondal\inst{1}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Brejesh Lall\inst{3}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Shiv Dutt Joshi\inst{3}\orcidID{xxxx-xxxx-xxxx-xxxx}\and Sumantra Dutta Roy\inst{3}\orcidID{xxxx-xxxx-xxxx-xxxx}}
\authorrunning{B. Das et al.}
\institute{Indian Institute of Technology Delhi, New Delhi - 110016, India \\ \email{\{bishshoy.das, milton.mondal, brejesh, sdjoshi, sumantra\}@ee.iitd.ac.in}}
\maketitle

\begin{abstract}
In standard neural network training, the gradients in the backward pass are determined
by the forward pass. As a result, the two stages are coupled. This is how most neural
networks are trained currently. However, gradient modification in the backward pass has
seldom been studied in the literature. In this paper we explore decoupled training,
where we alter the gradients in the backward pass. We propose a simple yet powerful
method called PowerGrad Transform (PGT), that alters the gradients before the weight
update in the backward pass and significantly enhances the predictive performance of the
neural network. PGT trains the network to arrive at a better optima at convergence. It
is computationally extremely efficient, virtually adding no additional cost to either
memory or compute, but results in improved final accuracies on both the training and
test sets. PGT is easy to integrate into existing training routines, requiring just a
few lines of code. PGT accelerates training and makes it possible for the network to
better fit the training data. With decoupled training, PGT improves baseline accuracies
for ResNet-50 by 0.73\%, for SE-ResNet-50 by 0.66\% and by more than 1.0\% for the
non-normalized ResNet-18 network on the ImageNet classification task.
\keywords{backpropagation \and gradients \and neural networks \and softmax \and clipping.}
\end{abstract}

@@

\begin{equation}
% \hspace{4cm}
W_i^{t+1} = W_i^t - \lambda\ \nabla_{W_i}(L)
\hspace{2cm}
i=D,D-1,\ldots,1
\end{equation}

@@

\begin{equation}
% \hspace{4cm}
W_i^{t+1} = W_i^t - \lambda\ f(\nabla_{W_i}(L))
\hspace{2cm}
i=D,D-1,\ldots,1
\end{equation}

@@

\begin{align}
% \hspace{4cm}
W_D^{t+1} &= W_D^t - \lambda\ f(\nabla_{W_D}(L)) & \\
W_i^{t+1} &= W_i^t - \lambda\ \nabla_{W_i}(L)^{*} &\hspace{-1cm} i=D-1,\ldots,1
\end{align}

@@

Gradient Clipping (GC): Gradient Clipping \cite{pascanu2013difficulty}, often used in
natural language processing methods \cite{merity2017regularizing}, is a technique that
involves changing or clipping the gradients with respect to a predefined threshold value
during backward propagation through the network and updating the weights using the
clipped gradients \cite{zhang2019gradient} \cite{smith2020generalization}. By rescaling the
gradients, the weight updates are likewise rescaled, significantly reducing the risk of
an overflow or underflow \cite{pascanu2012understanding}. GC can be used for training
networks without batch-normalization.

@@

Knowledge Distillation: The formulation of PowerGrad Transform is akin to temperature
based logit suppression of \cite{hinton2015distilling}. Knowledge distillation
\cite{DBLP:conf/nips/BaC14} is a process in which two networks are trained. First a
teacher network is trained on a given dataset and then the soft-labels (the predicted
probabilities) of the teacher network are used to train another network, called the
student network. In similar style of label smoothing, knowledge distillation aims to
generate smooth gradient behaviour by forcing the network not to overpredict on any
single image. As in case of label smoothing, the student network's weights are
automatically penalized if the network assigns a probability value higher than the soft
labels generated by the teacher network. Variants of knowledge distillation include
self-distillation \cite{zhang2019your} \cite{wang2021memory}, identical student network
distillation \cite{furlanello2018born}, channel distillation \cite{ge2019distilling},
regularizing wrong predictions \cite{yun2020regularizing} and representation or
embedding based knowledge distillation \cite{aguilar2020knowledge} \cite{yao2018deep}
\cite{passalis2018unsupervised}. Distillation is applied to various tasks such as model
compression \cite{wang2020real}, face recognition \cite{ge2018low}, enhancing ensemble
performance \cite{zhang2018deep}, interpreting neural networks \cite{liu2018improving}.
Efficacy of distillation is extensivly studied in \cite{cho2019efficacy}
\cite{yuan2020revisiting}. Applications of logit suppression include: deep metric
learning \cite{zhai2018classification}, non-parametric neighborhood component analysis
\cite{wu2018improving}, sequence-to-sequence modelling \cite{chorowski2016towards},
reinforcement learning \cite{he2018determining} and face recognition
\cite{liu2017sphereface} \cite{wang2018cosface}.

@@

\begin{proof} \begin{alignat}{2} &\ \ \frac{p_i^{\alpha }}{\sum
_{j=1}^C p_j^{\alpha}} < p_i \\ \Rightarrow &\ p_i^{\alpha -1}<\sum _{j=1}^C p_j^{\alpha
} \\ \Rightarrow &\ p_i > \Big(\sum _{j=1}^C p_j^{\alpha}\Big)^{\frac{1}{\alpha-1}}
\end{alignat} \end{proof}

@@

\begin{proof} \begin{equation} \frac{\partial}{\partial \alpha
}\left(\sum _{j=1}^c p_j^{\alpha}\right){}^{\frac{1}{\alpha -1}}\ \ \geq \ \ 0
\end{equation} \end{proof}

@@

\begin{table}[!t]
\centering
\caption{ Results and comparison table for CNN networks trained on ImageNet-1K (sorted
in descending order w.r.t. the last column denoting test accuracy improvements over
baseline). Best training and test accuracies are highlighted in red and blue
respectively. Accuracy differences are highlighted in yellow. In all experiments, we
observe that PGT positively improves both training and test accuracies. We also observe
that deeper networks such as SE-ResNet-50 and ResNet-50 accrue higher gains over the
baselines as compared to shallow networks like ResNet-18. }
\label{tab:imagenet_table}
\begin{tabular}{cccccccc}
\multirow{2}{*}{\textbf{Model}} & \multirow{2}{*}{\textbf{Scheduler}} &
\multirow{2}{*}{\textbf{Method}} & \textbf{PGT} & \textbf{\hspace{-0.65cm} Train}
& \textbf{Train} & \textbf{\hspace{-0.65cm} Test} & \textbf{Test} \\
& & & \textbf{($\alpha$)} & \textbf{Acc. (\%)} & \textbf{Diff (\%)} &
\textbf{Acc. (\%)} & \textbf{Diff (\%)} \\
\midrule
\multirow{2}{*}{SE-ResNet-50} & \multirow{2}{*}{Cosine} & Baseline & - & 81.5 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.97}}} & 77.218 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.734}}} \\
& & PGT & 0.3 & \textcolor{red}{\textbf{82.47}} & & \textcolor{blue}{\textbf{77.952}} &
\\
\midrule
\multirow{2}{*}{ResNet-50} & \multirow{2}{*}{Cosine} & Baseline & - & 79.18 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.5}}} & 76.56 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.656}}} \\
& & PGT & 0.05 & \textcolor{red}{\textbf{79.68}} & & \textcolor{blue}{\textbf{77.216}} &
\\
\midrule
\multirow{2}{*}{ResNet-50} & \multirow{2}{*}{Step} & Baseline & - & 78.99 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.57}}} & 75.97 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.524}}} \\
& & PGT & 0.05 & \textcolor{red}{\textbf{79.56}} & & \textcolor{blue}{\textbf{76.494}} &
\\
\midrule
\multirow{2}{*}{ResNet-101} & \multirow{2}{*}{Cosine} & Baseline & - & 82.29 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.81}}} & 77.896 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.362}}} \\
& & PGT & 0.3 & \textcolor{red}{\textbf{83.1}} & & \textcolor{blue}{\textbf{78.258}} &
\\
\midrule
\multirow{2}{*}{SE-ResNet-18} & \multirow{2}{*}{Cosine} & Baseline & - & 71.42 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.18}}} & 71.09 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.346}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{71.6}} & & \textcolor{blue}{\textbf{71.436}} &
\\
\midrule
\multirow{2}{*}{ResNet-18} & \multirow{2}{*}{Step} & Baseline & - & 69.95 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.35}}} & 69.704 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.14}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{70.3}} & & \textcolor{blue}{\textbf{69.844}} &
\\
\midrule
\multirow{2}{*}{ResNet-18} & \multirow{2}{*}{Cosine} & Baseline & - & 70.38 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.15}}} & 70.208 &
\textcolor{olive}{\multirow{2}{*}{\textbf{+0.09}}} \\
& & PGT & 0.25 & \textcolor{red}{\textbf{70.53}} & & \textcolor{blue}{\textbf{70.298}} &
\\
\end{tabular}
\end{table}

@@

In our experiments with Squeeze-and-Excitation variant of ResNet-50 i.e.
SE-ResNet-50\cite{hu2018squeeze}, we observe significant improvements; a $0.97\%$ boost
in training accuracy and a $0.734\%$ increase in test accuracy. In our experiments with
ResNet-50, we find an increase of $0.5\%$ and $0.656\%$ performance enhancement over the
cosine scheduler baseline for training and testing respectively. The corresponding
improvement over the step scheduler baseline for ResNet-50 is $0.57\%$ (training) and
$0.524\%$ (testing). ResNet-101 sees a higher improvement in training fit, $0.81\%$ to
be exact, while the improvement over the test set is $0.362\%$. Smaller networks such as
SE-ResNet-18 and ResNet-18 sees accuracy boosts which are smaller but nevertheless
positive. Therefore we conclude that with PGT, the consistent improvements in training
accuracies across all cases is because the networks train better and arrive at better
optimas during convergence. Per epoch training and test accuracy plots of ResNet-18 and
ResNet-50 (both with and without PGT) are shown in Fig. \ref{fig:metrics}(a, b). For
better visualization of the top end of the plots, we scale both axes in log scale.
Because both training and test accuracies improve, it indicates that PowerGrad Transform
allows the network to learn better representations\footnote{Reproducible code, training
recipes for the experiments, pretrained checkpoints and training logs are provided at:
\url{ https://github.com/bishshoy/power-grad-transform }. Code is adapted from
\cite{rw2019timm}.}.

@@

\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{baseline-w-layer-4-2}
\caption{Norm of filters of convolutional layer 11 (without PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{baseline-w-layer-7-2}
\caption{Norm of filters of the final convolutional layer (without PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{baseline-f-layer-22-1}
\caption{Norm of final layer's output features (without PGT)}
\end{subfigure}

\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{pgt-w-layer-4-2}
\caption{Norm of filters of convolutional layer 11 (with PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{pgt-w-layer-7-2}
\caption{Norm of filters of the final convolutional layer (with PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{pgt-f-layer-22-1}
\caption{Norm of final layer's output features (with PGT)}
\end{subfigure}
\captionsetup{font=normalsize}
\caption{ Experiment demonstrating the Zeroing Out phenomena, as observed in different
filters and feature vectors. PGT greatly reduces the zeroing out of filters and promote
better learning. }
\label{fig:norm_plots}
\end{figure}

@@

\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{baseline-w-layer-5-3}
\caption{Norm of filters of layer 15 (without AGC/PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{pgt-w-layer-5-3}
\caption{Norm of filters of layer 15 (with PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{agc_pgt-w-layer-5-3}
\caption{Norm of filters of layer 15 (with AGC+PGT)}
\end{subfigure}

\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{baseline-w-layer-1-2}
\caption{Norm of filters of layer 2 (without AGC/PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{pgt-w-layer-1-2}
\caption{Norm of filters of layer 2 (with PGT)}
\end{subfigure}
\begin{subfigure}[t]{0.32\textwidth}
\includegraphics[width=\textwidth]{agc_pgt-w-layer-1-2}
\caption{Norm of filters of layer 2 (with AGC+PGT)}
\end{subfigure}
\captionsetup{font=normalsize}
\caption{ Experiment demonstrating the efficacy of PGT and AGC+PGT, as observed in
different layers. }
\label{fig:agc_pgt}
\end{figure}

@@

\bibliographystyle{splncs04}
\bibliography{references}
